Functional diversity improves tropical forest resilience to disturbance

### 2018-04-23 draft.Rmd
Draft 0.1.9004 Third version for IM review

---
title: Functional diversity improves tropical forest resilience to disturbance
author: "Sylvain Schmitt, Bruno Hérault, et al"
date: '`r Sys.Date()`'
output:
  github_document: default
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
  bookdown::word_document2: default
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
    keep_tex: true
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

Natural disturbance trough treefall is a main factor of forest dynamic [@Veblen1992]. Still, climate and human activities are also at the origin of forest disturbances, with huge and underevaluated impacts. Human-induced forest disturbances will result either in forest degradation or in deforestation. Degradation has met numerous definitions [@Simula2009], but can be defined as the result from a disturbance event that induced a modification of the forest ecosystem reducing ecosystem services while keeping the ecosystem as a forest, contrary to deforestation. Degradation can be the result of forest logging and management, whereas deforestation can be realized through total afforestation or fragmentation resulting in land use change. @Pearson2017 encompassed selective timber harvest in native forests, woodfuel harvest and fire in tropical forest degradation. Tropical forests disturbances, through deforestation and degradation, account for 8.26 billion of tons of carbon dioxide emissions per year [@Pearson2017]. It represents the third source of greenhouse gas. Besides deforestation has been rightly the focus of worldwide attention, degradation has been less studied and quantified through tropics. Degradation from forest has been estimated to represents 10 time the deforestation area [@Herold2011] and represents 20% of greenhouse gas emissions in the Brazilian Amazon [@Asner2005]. In addition to the direct impact of human on forests, a more insidious disturbance is climate change [@Herault2016].

Regarding deforestation, the ecosystem response to land use change is almost absent due to the negative feedback of the new ecosystem state after the tipping point was reached. Consequently, ecosystem response to disturbance should be studied in the different contexts of forest degradation. Concerning timber and woodfuel harvest, @DeAvila2015 found that additional sylvicultural practices and logging damages had stronger negative effect on diversity than logging itself. @Carreno-Rocabado2012 found no direct effect of medium intensity logging on species diversity but a community drift towards 'fast' acquisitive species [@reich_world-wide_2014]. However, even the short term impact on biodiversity is unclear, as sometimes short term increases in diversity can occur at low sylvicultural intensities [@Martin2015]. Forest fires have been shown to increase pioneers abundance, and if recurrent, to impoverish the forests in the eastern Amazon [@Cochrane1999a]. Finally, climate change effect on tropical areas results in a significant risk of precipitation decrease [@Davidson2012]. In addition, intense drought events has been shown to result in huge mortality rates and biomass loss [@Phillips2009].

Tropical rainforests have fascinated ecologists due to their outstanding diversity [@connell_diversity_1978]. Effectively tropical forests host over half of the Earth's biodiversity [@Scheffers2012]. High biodiversity from tropical rainforests is the source of many ecosystem functions. Among others, tropical forests play a key role in biogeochemical cycles, including carbon storage [@Lewis2004]. Ecosystem functions from tropical forests support numerous ecosystem services, such as timber production and climate regulation.

The question of disturbance impact on tropical forest can be directly related to the emerging field of biodiversity and ecosystem functioning [@Loreau2000]. Tropical forest outstanding biodiversity will be both a factor and a result of forest ecosystem response to disturbance. And forest ecosystem response to disturbance will directly modify ecosystem functioning in both short and long term. Consequently assessing disturbance effect on tropical forest linking diversity and ecosystem functioning seems an obvious and promising way [@Loreau2010].

Negative short term impacts of disturbances have been assessed for logging, fires and droughts ; but much less is known about the long term impact [@Osazuwa-Peters2015]. The main reason is the difficulty to conduct long term empirical study [but see @Herault2010], which can be completed by the use of forest simulators [@Huth2004; @Khler2004; @Ruger2008; @Tietjen2006]. Individual-based models of forest dynamics present the perfect framework to develop such joint biodiversity-ecosystem approaches [@Marechaux2017]. Individual-based models describe forest accumulating carbon through time, assessing tree growth, or releasing carbon through gap opening [@Bugmann2001]. Up to several dozens of different Plant Functional Types (PFTs) are generally defined and models can sometimes be fully spatially explicit [@Pacala1996]. Recently, the forest growth simulator TROLL [@Chave1999], an individual-based and spatially explicit forest model, was developed to introduce recent advances in plant ecophysiology. TROLL model relates physiological processes to species-specific functional traits [@Marechaux2017]. Consequently, TROLL model allow to simulate fully a neotropical forest biodiversity to study biodiversity-ecosystem functioning link response to a disturbance.

We decided to use the forest model TROLL to study the role of biodiversity in forest ecosystem answer to disturbance. Resilience encompass several definitions but was summarized by @Oliver2015 as the degree of resistance or fast recovery from an ecosystem function to environmental disturbance.

Our work is based on the general hypothesis of a positive relationships between biodiversity and productivity. We assumed that when a disturbance event happen, due to a higher productivity, forest with an increased diversity will recover quicker and thus be more resilient, besides other mechanisms could be involved in an increased resilience from diverse forest. This theoretical expectation stand on two processes: complementarity and selection effects [@Loreau2001]. If we take a species pool with different productivity in monoculture, their assemblage will allow them an overall higher productivity due to a better resources acquisition through resources partitioning and niche differentiation. Additionally, some species will individually have an higher productivity in the assemblage than in monoculture due to facilitation from other species present in the assemblage. Complementarity effect is the addition of the better resource acquisition and the facilitation. Now if we look at the evolution of the species assemblage over time, more competitive species will progressively dominate through competitive selection. And if competitive species are more productive, they will increase assemblage overall productivity. This resources preemption by more competitive species is the selection effect. Moreover, we expect the sampling effect to improve resilience of species-rich forests. A bigger initial sampling of the regional species pool in a rich forest, will allow more redundancy and a lower risk to lose important functional traits assemblages for the ecosystem when the disturbance happen.

@@

We suggest to use a process-based model to overcome those limitations because TROLL model have the same key features that @Morin2011: (i) TROLL has been developed and validated independently from the biodiversity and ecosystem functioning issue, (ii) TROLL use 163 real tropical tree species characterized by their functional traits, (iii) TROLL allows to study long term effects with realized species richness for several centuries, (iv) TROLL allows to explore more species associations than real experiments, (v) TROLL allows to explore process resulting in observed relations. In addition, TROLL has been designed to be used in species-rich ecosystems such as tropical rainforests.

In the present study, we focused on mechanisms involved in the relationship between biodiversity and forest ecosystem resilience. We used a simulation approach using TROLL model to assess long term processes for different levels of disturbances. We first assessed diversity effect on forest resilience of structure and functioning using numerous indices of taxonomic and functional diversities. Then, we measured biodiversity net effect, partitioned into complementarity and selection effects, on resilience for several ecosystem metrics.

@@

## Model

TROLL model each tree individually in a located environment. Thus TROLL model, alongside with SORTIE  [@Pacala1996; @Uriarte2009] and FORMIND [@Fischer2016; @Kohler1998], can be defined as an individual-based and spatially explicit forest growth model. TROLL simulates the life cycle of individual trees from recruitment, with a diameter at breast height (dbh) above 1 cm, to death with growth and seed production. Trees are growing in a spatialized light environment explicitly computed within voxels of 1 $m^3$. Each tree is consistently defined by its age, diameter at breast height (dbh), height (h), crown radius (CR), crown depth (CD) and leaf area (LA). Tree geometry is calculated with allometric equations but leaf area vary dynamically within each crown following carbon allocations. Voxels resolution of 1 $m^3$ allow the establishment of maximum one tree by 1x1 m pixels. Each tree is flagged with a species label inherited from the parent tree through the seedling recruitment. A species label is associated to a number of species specific parameters (see table \@ref(tab:traits)) related to functional trait values which can be sampled on the field. Carbon assimilation is computed over half-hourly period of a representative day. Then allocation is computed to simulate tree growth from an explicit carbon balance (in contrast to previous models). Finally environment is updated at each time step set to one month. Seedlings are not simulated explicitly but as a pool. In addition below-ground processes, herbaceous plants, epiphytes and lianas are not simulated inside TROLL. The source code is written in C++ and available upon request. All modules of TROLL models are further detailed in @Marechaux2017.

@@

Previous implementation of TROLL model used @Reich1991a allometry to infer leaf lifespan $LL$ from species leaf mass per area $LMA$ [@Marechaux2017]. But the use of the allometrie from @Reich1991a with current implementation of the TROLL model resulted in an underestimation of leaf lifespan for low LMA species. Consequently we designed our own leaf lifespan allometry, specific to tropical tree species, to be used in subsequent analysis (see supplementary materials S1).

We decided to explore the resilience of forest ecosystem following a generic disturbance. Disturbance module inside TROLL was designed in the simplest way in order to relate the ecosystem answer to volume loss without any individuals nor species targeting. For a given iteration $disturb_{iter}$, individuals are picked randomly with a uniform law on the number of trees. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined lost basal area ($disturb_{intensity}$ in % of BA).

@@

In order to assess the role of biodiversity in ecosystem answer to both disturbance and sylviculture, we needed to create a space of experiments encompassing both variation of disturbance and biodiversity. Disturbance was represented by percentage of basal area loss (0%, 25%, 50% and 75%), using default parameters. Biodiversity was integrated with two of its components: taxonomic and functional diversities. We used species richness $SR$ to represents taxonomic diversity (5, 25, and 125 species). Functional diversity can be related to numerous components, and @Borgy2017 argued for 5: richness, divergence, regularity, overlap and mean. Because mature forest were created from a bare soil with TROLL simulations, we could not control a priori divergence, regularity and overlap but only assess them after running the simulations, i.e. before applying the disturbance. Consequently, we focused on functional richness with convex hull volume $CHV$ and functional mean with community weighted mean $CWM$. For each level of species richness $SR$, we selected 20 communities with growing convex hull volume $CHV$ but with a community weighted means close to the regional species pool community weighted means. Effectively, we did not wanted drastic change in community means that could have more effect than functional richness itself. This design of experiments resulted in 60 communities ($3~SR*20~CHV$) and 240 simulations ($60 ~communities*4~levels~of~disturbance$) over 600 years (maturity being assumed after 500 years of regeneration [@Marechaux2017]). Functional diversities of mature forests were assessed with @villeger_new_2008 indices (FRIC, FEve, FDiv, and FDis).

## Ecosystem response analysis

### Ecosystem functions

Tropical forest ecosystems provides numerous ecosystem services linked to several ecosystem functions. We decided to describe simulated tropical forests in two major functions: forest structure and forest functioning. Forest structure was represented by aboveground biomass ($AGB$ in $ton~C.ha^{-1}$), basal area ($BA$ in $m^2.ha^{-1}$), total number of stem ($N$), number of stem above $10~cm$ diameter ($N10$), and number of stem above $30~cm$ diameter ($N30$). Forest functioning was represented by growth primary productivity ($GPP$ in $MgC.ha^{-1}$), net primary productivity ($NPP$ in $MgC.ha^{-1}$), tree autotrophic respiration in day ($Rday$ in $MgC.ha^{-1}$) and tree autotrophic respiration in night ($Rnight$ in $MgC.ha^{-1}$). Variables representing both forest functioning and forest structure are sometimes highly correlated within each group. But multiple variables allow a better assessment of ecosystem equilibrium.

The resilience of metrics values post disturbance were assessed through @Henry2012 formula:

@@

The resilience of the system $R(t)$ at the time $t$ is described by the ratio of recovery $Recovery(t)$ at time $t$ to loss suffered $Loss(t_d)$ at disturbance time $t_d$. But in our peculiar case of tropical forest ecosystems, the equilibrium used to calculate $Loss(t_d)$ can not be reduced to a specific time if the equilibrium is dynamic. Consequently, to encompass undisturbed ecosystem variations through time, we simulated an undisturbed control ecosystem $C$. And the resilience of the system $R(t)$ at the time $t$ was defined as the ratio of the ecosystem metric values in the disturbed simulation $X_T(t)$ over the ecosystem metric values from the control $X_C(t)$ (the control simulations being relatively stable due to a small stochasticity among TROLL simulations [@Marechaux2017]). Thus, the value of resilience $R(t)$ is normalized for all simulations and metrics. $R(t)$  will be equal to $R_{eq} = 1$ when reaching the equilibrium value. Consequently we can calculate an euclidean distance to equilibrium $d_{eq}(t)$ as $d_{eq}(t) = \sqrt{(R_{eq} - R(t))^2}$. Ecosystem euclidean distance to equilibrium was calculated in a multi-dimensional space for the two functions described above: forest structure (AGB, BA, N, N10, and N30) and forest functioning (GPP, NPP, Rday, and Rnight). We then used integrated euclidean distance to equilibrium over time to assess simulations resilience.

We transformed all ecosystem outputs from the 240 disturbance simulations in resilience metrics normalizing the treatment values by their corresponding control (Figure \@ref(fig:datatrans) A and B). We then gathered ecosystem outputs by main ecosystem functions (forest structure and forest functioning) to compute ecosystem distance to equilibrium (Figure \@ref(fig:datatrans) C). Finally, we integrated distance to equilibrium in a cumulative sum over time (Figure \@ref(fig:datatrans) D). The ranking of simulations regarding disturbance intensity was stable over time for the 240 simulations. So we used the cumulative integral after 600 years $Ieq_{600}$ as a measurement of ecosystem resilience.

```{r datatrans, echo=FALSE, fig.cap='Ecosystem outputs data transformation. Ecosystem outputs (**A**) are normalized by the control value over time to calculate resilience (**B**); resilience of different ecosystem outputs is then used in a multidimensional space to caluclate ecosystem distance to equilibrium (**C**); finally distance ot equilibirum is integrated over time in a cummulative sum (**D**). ', cache=TRUE, fig.height=8, fig.width=12, message=FALSE}
cols <- RColorBrewer::brewer.pal(3, "YlOrRd")
g_agb <- ecosystem %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = agb/1000, color = disturbance)) +
  geom_point() +
  xlab('Time (years)') +
  ylab('Aboveground biomass (tonC/ha)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                       values=c("#999999", cols))
g_resilience <- resilience %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = agb, color = disturbance)) +
  geom_hline(yintercept = 1, col = 'red') +
  geom_point() +
  xlab('Time (years)') +
  ylab('Resilience of aboveground biomass') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
g_deq <- deq %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = structure, 
             color = disturbance, fill = disturbance)) +
  geom_hline(yintercept = 0, col = 'red') +
  geom_point() +
  geom_area(alpha = 0.2) +
  xlab('Time (years)') +
  ylab('Distance to forest structure equilibrium\n(AGB, BA, N, N10 and N30)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols) +
  scale_fill_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
g_Ieq <- Ieq %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = Ieq.structure, color = disturbance)) +
  geom_point() +
  xlab('Time (years)') +
  ylab('Cummulative integral of\ndistance to forest structure equilibrium\n(AGB, BA, N, N10 and N30)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
cowplot::plot_grid(g_agb, g_resilience, g_deq, g_Ieq, labels = LETTERS[1:4])
```

@@

Biodiversity is not only a facet of the experimental design and an ecosystem output through forest diversity, but also interact on ecosystem functioning and consequently on its answer to disturbance. Biodiversity ecosystem functioning relation can be split in complementarity and selection effect with @Loreau2001 partitioning:

@@

$X_{sp}$ is the ecosystem variable value for one species either in mixture $X_{sp}(mixture)$ or in monoculture $X_{sp}(monoculture)$. $P_{sp}$ is the proportion of the species in the mixture represented by species relative abundance. Consequently, $CE$ averages diversity effects of all species presents in the mixture (both negatives and positives). Whereas $SE$ become positive when dominant species outperform themselves in mixture than in monoculture, and negative when less dominant species outperform themselves in mixture than in monoculture [@Tobner2016]. But similarly to resilience measurement, biodiversity net effect $NE$ is a dynamic equilibrium and vary over time without disturbance. So in order to correctly assess selection and complementarity effect in answer to disturbance, we normalized it by undisturbed control ecosystem net effect $NE_C$ to measure treatments net effect resilience $R(NE_T)$:

@@

Resilience trajectories of ecosystem variable after disturbance were partitioned between complementarity effect $CE$ and selection effect $SE$. In order to do that, the design of experiment was repeated for the each of the 163 species individually representing 652 simulations of monoculture.

@@

We compared cumulative integral after 600 years to communities taxonomic and functional diversity for each level of disturbance (see Figure \@ref(fig:gIeq)). We found that increased functional diversity [FDiv, @villeger_new_2008] was reducing cumulative integral from ecosystem distance to forest structure equilibrium after 600 years ($Ieq_{600}$). In addition, functional evenness was complementary reducing $Ieq_{600}$. Finally species richness was not directly link to $Ieq_{600}$. Effectively, low species richness could result in variant $Ieq_{600}$, but increased species richness resulted in increased functional diversity and consequently lower $Ieq_{600}$. We found similar results for all disturbance levels and forest functioning (see [Appendix 5: Disturbance simulations]).

@@

We measured all ecosystem outputs biodiversity net effect in disturbance simulations by comparing them to their species corresponding monoculture simulations. The net effect was then partition between selection and complementarity effect. We normalized complementarity and selection effect of disturbed simulations by biodiversity net effect of undisturbed control simulation (see Table \@ref(tab:tNE)) to measure the resilience of each ecosystem outputs, e.g. aboveground biomass (see Figure \@ref(fig:gBE)). We found that complementarity effect was recovering biodiversity net effect on aboverground biomass and was stronger than selection effect in the first decades. But after few decades the complementarity effect diminished toward a low value. On the contrary selection effect was reduced or even removed by the disturbance. It increased during the whole simulation and was greater than complementarity effect only after decades. The time lag for which complementarity effect was greater than selection effect was increasing with disturbance intensity. Finally 600 years after the disturbance event biodiversity net effect on aboveground biomass was still not recovered for a disturbance intensity greater than 25% of basal area. We found similar results but with an amplified signal for basal area ($BA$) and stem abundance ($N$ but with an inverted signal because of the forest self thinning) (see Supplementary materials S2). Finally, forest growth primary productivity ($GPP$) recovered in few years (proportionally to disturbance), and its net effect was maintained by complementarity effect (see Supplementary materials S2).

@@

In the limit of the model, we were able to show that diversity improved tropical forest resilience. More particularly, functional diversity and evenness are key components of diversity in forest recovery after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in forest succession start with facilitation before more productive species dominate the forest and insure recovery. We thus advocate the view that both diversity theory (with niche complementarity and facilitation) and mass ratio theory (represented by selection) represents the mechanisms at the origin of functional diversity effect on ecosystem functions depending on the spatial scale and the time since the disturbance event.

## Diversity improve tropical forest resilience

>>>>

---
title: Functional diversity improves tropical forest resilience to disturbance
author: "Sylvain Schmitt, Isabelle Maréchaux, Jerome Chave, Fabien Fischer, Camille Piponiot, Stéphane Traissac & Bruno Hérault"
date: '`r Sys.Date()`'
output:
  bookdown::html_document2:
    number_sections: false
    toc: true
    toc_float: yes
  github_document: default
  bookdown::pdf_document2:
    includes:
      before_body: ./template/doc_prefix.tex
      in_header: ./template/preamble.tex
    number_sections: false
    toc: false
    keep_tex: true
  bookdown::word_document2: default
linestretch: 1.5
csl: /home/sylvain/Documents/Bibliography/csl/mee.csl
bibliography: /home/sylvain/Documents/Bibliography/library.bib
link-citations: yes
---

@@

Natural disturbances caused by the death of one or more trees, are the dominant forms of forest renewal in tropical forests as the creation of canopy gaps continuously modifies forest horizontal and vertical structure [@Veblen1992]. Today, climate change and other direct human activities are modifying forest disturbance regimes, with huge and under evaluated impacts. When human induced disturbances are high, it is referred to as "degraded forests" characterized by a modification of the forest ecosystem functioning, possibly reducing the furnished ecosystem services, while keeping the ecosystem as a forest, contrary to deforestation [@Simula2009]. Besides deforestation has been rightly the focus of worldwide attention, degradation has been less studied and quantified through tropics. The main source of degradation around the tropics is forest logging that encompass selective timber harvest, in native forests, and wood fuel harvest, often in secondary forests [@Pearson2017]. Tropical forests disturbances account for 2.07 billion of tons of carbon dioxide emissions per year [@Pearson2017], representing 20% of greenhouse gas emissions in the Brazilian Amazon [@Asner2005], estimating to cover areas 10 or more times larger than deforestation itself [@Herold2011]. Finally, disturbance in tropical forest dynamics induced by climate change likely results from significant changes in precipitation regime [@Davidson2012] with intense drought events resulting in huge mortality rates and biomass loss [@Phillips2009].

Tropical forest ecosystem response to disturbance depends first on the disturbance intensity itself. For instance, @DeAvila2015, focusing on timber and wood fuel harvest in Central Amazonia, found that when disturbance is high, e.g. induced by post logging sylvicultural practices had stronger negative effect on the post-disturbance system trajectory than logging itself. @Carreno-Rocabado2012 found no direct effect of disturbance intensity logging on community diversity itself but a community composition drift towards 'fast' acquisitive species [@reich_world-wide_2014]. However, even the short-term impact on biodiversity is unclear, as sometimes short term increase in diversity can occur at low disturbance intensities [@Martin2015].

The main reason of our lack of knowledge is the difficulty to conduct long term empirical study so that the use of forest simulators has become inevitable [@Huth2004]. Individual-based models of forest dynamics with species-explicit model parameters present the perfect framework to develop such joint biodiversity-ecosystem functioning studies [@Marechaux2017]. Starting from scratch, these models allow describing forest recovery through time, assessing individual tree recruitment, growth and mortality [@Bugmann2001]. Managing diversity in such models is still challenging and many models have built open Plant Functional Types to manage the volume of model parameters. [@Pacala1996].

Recently the forest growth simulator TROLL [@Chave1999], an individual-based and spatially-explicit forest model has been explicitly developed to deal with high diversity levels by using plant ecophysiological traits as surrogates of model parameters [@Marechaux2017]. Consequently, TROLL allow to fully simulate a highly-diverse system and offers the opportunity to study biodiversity-ecosystem functioning *in silico*.

Our work is based on the general hypothesis of a positive relationships between biodiversity and productivity. We assumed that when a disturbance event happen, due to a higher productivity, more diverse forests will recover quicker and thus be more resilient. This theoretical expectation stand on two processes: complementarity and selection effects [@Loreau2001]. Complementarity effect is the addition of better resources acquisition, due to resources partitioning and niche differentiation, and facilitation in diverse communities. Selection effect results from resources preemption by more competitive species. Moreover, we expect the sampling effect to improve resilience of species-rich forest, with redundancy and a lower risk to lose important functional traits.

@@

In the present study, we thus focused on mechanisms involved in the relationship between biodiversity and forest ecosystem resilience. We used a simulation approach using TROLL model to assess long term processes for different levels of disturbances. We first assessed diversity effect on forest resilience of structure and functioning using numerous indices of taxonomic and functional diversities. Then, we measured biodiversity net effect, partitioned into complementarity and selection effects, on resilience of simulated tropical forest communities to disturbance.

@@

## TROLL Model

TROLL is an individual-based and spatially explicit forest growth model. TROLL simulates the life cycle of individual trees from recruitment, with a diameter at breast height (dbh) above 1 cm, to death with growth and seed production. Trees are growing in a spatialized light environment explicitly computed within voxels of 1 $m^3$. Each tree is consistently defined by its age, diameter at breast height (dbh), height (h), crown radius (CR), crown depth (CD) and leaf area (LA). Tree geometry is calculated with allometric equations but leaf area vary dynamically within each crown following carbon allocations. Voxels resolution of 1 $m^3$ allow the establishment of maximum one tree by 1x1 m pixels. Each tree is flagged with a species label inherited from the parent tree through the seedling recruitment. A species label is associated to a number of species specific parameters (Table \@ref(tab:traits)) related to functional trait values. Carbon assimilation is computed over half-hourly period of a representative day. Then allocation is computed to simulate tree growth from an explicit carbon balance. Environment is updated at each time step set to one month. Seed-rains and seedlings are not simulated explicitly but as a pool. In addition below-ground processes, herbaceous plants, epiphytes and lianas are not simulated inside TROLL. The source code is written in C++ and available upon request. All modules of TROLL models are further detailed in @Marechaux2017. Following an underestimation of leaf lifespan for low LMA species, we designed our own leaf lifespan allometry, specific to tropical tree species, to be used in subsequent analysis (see supplementary materials S1).

@@

The disturbance module inside TROLL was designed in the simplest way by, for a given iteration $disturb_{iter}$, picking individual trees randomly up with a uniform law and without any individuals nor species targeting. Selected individuals are then removed without triggering a treefall to avoid any side effect. The operation is repeated until the disturbance result in a defined lost basal area ($disturb_{intensity}$).

@@

In order to assess the role of biodiversity in ecosystem response to disturbance, we  simulated post-disturbance forest trajectories, encompassing both variation of disturbance intensity and biodiversity levels. The gradient of disturbance intensity was represented by percentage of basal area loss (0%, 25%, 50% and 75%). Biodiversity was assessed with two perspectives: taxonomic and functional. We used species richness $SR$ to simulate initial assemblages with diverse taxonomic diversity (5, 25, and 125 species). Functional diversity was assessed after running the initial simulations, *i.e.* before applying the disturbance. We focused on functional richness with convex hull volume $CHV$ and functional mean with community weighted mean $CWM$. For each level of species richness $SR$, we simulated 20 communities with a large range of convex hull volume $CHV$ but with a community weighted means close to the regional species pool community weighted means. This design of experiments resulted in 60 tree communities ($3~SR*20~CHV$) and 240 simulations ($60 ~communities*4~levels~of~disturbance$) over 600 years (maturity being assumed after 500 years of regeneration [@Marechaux2017]). Functional diversities at the end of simulations were assessed with @villeger_new_2008 indices (FRIC, FEve, FDiv, and FDis).

## Analysing Ecosystem Response

### Ecosystem structure and functioning

We described the simulated tropical forests from changes in values of forest structure (aboveground biomass ($AGB$ in $ton~C.ha^{-1}$), basal area ($BA$ in $m^2.ha^{-1}$), total number of stem ($N$), number of stem above $10~cm$ diameter ($N10$), and number of stem above $30~cm$ diameter ($N30$)) and in values of forest functioning (growth primary productivity ($GPP$ in $MgC.ha^{-1}$), net primary productivity ($NPP$ in $MgC.ha^{-1}$), tree autotrophic respiration in day ($Rday$ in $MgC.ha^{-1}$) and tree autotrophic respiration in night ($Rnight$ in $MgC.ha^{-1}$)).

The resilience of forest structure and functioning values was assessed with:

@@

The resilience of the system $R(t)$ at the time $t$ is described by the ratio of recovery $Recovery(t)$ at time $t$ to loss suffered $Loss(t_d)$ at disturbance time $t_d$. But in our peculiar case of simulated forest ecosystems, the equilibrium used to calculate $Loss(t_d)$ is dynamic in time so that we simulated an undisturbed control ecosystem $C$. And the resilience of the system $R(t)$ at the time $t$ was defined as the ratio of the ecosystem metric values in the disturbed simulation $X_T(t)$ over the ecosystem metric values from the control $X_C(t)$ (the control simulations being relatively stable due to small stochasticity among TROLL simulations [@Marechaux2017]). $R(t)$  will be equal to $R_{eq} = 1$ when reaching the equilibrium value. Next, we calculated an euclidean distance to equilibrium $d_{eq}(t)$ as $d_{eq}(t) = \sqrt{(R_{eq} - R(t))^2}$. Ecosystem euclidean distance to equilibrium was calculated in a multi-dimensional space for (the two functions described above: (i) forest structure (AGB, BA, N, N10, and N30) and (ii) forest functioning (GPP, NPP, Rday, and Rnight). We then integrate the euclidean distances over time to get our ecosystem resilience indicators.

We then applied the following to each indicator:

1. Normalizing the treatment values by their corresponding control (Figure \@ref(fig:datatrans) A and B).
2. Respectively gathering forest structure and forest functioning together to get synthetic indicators (Figure \@ref(fig:datatrans) C)
3. Integration over time (Figure \@ref(fig:datatrans) D)
4. The ranking of simulations regarding disturbance intensity was highly stable over time for the 240 simulations. So we used the cumulative integral after 600 years $Ieq_{600}$ as a measurement of ecosystem resilience.

```{r datatrans, echo=FALSE, fig.cap='Ecosystem outputs data transformation. Ecosystem outputs (**A**) are normalized by the control value over time to calculate resilience (**B**); resilience of different ecosystem outputs is then used in a multidimensional space to caluclate ecosystem distance to equilibrium (**C**); finally distance at equilibrium is integrated over time in a cumulative sum (**D**). ', cache=TRUE, fig.height=8, fig.width=12, message=FALSE}
cols <- RColorBrewer::brewer.pal(3, "YlOrRd")
g_agb <- ecosystem %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = agb/1000, color = disturbance)) +
  geom_point() +
  xlab('Time (years)') +
  ylab('Aboveground biomass (tonC/ha)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                       values=c("#999999", cols))
g_resilience <- resilience %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = agb, color = disturbance)) +
  geom_hline(yintercept = 1, col = 'red') +
  geom_point() +
  xlab('Time (years)') +
  ylab('Resilience of aboveground biomass') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
g_deq <- deq %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = structure, 
             color = disturbance, fill = disturbance)) +
  geom_hline(yintercept = 0, col = 'red') +
  geom_point() +
  geom_area(alpha = 0.2) +
  xlab('Time (years)') +
  ylab('Distance to forest structure equilibrium\n(AGB, BA, N, N10 and N30)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols) +
  scale_fill_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
g_Ieq <- Ieq %>% 
  left_join(., doe, by = 'sim') %>%
  filter(richness == 125) %>% 
  filter(cvh == 1) %>% 
  ggplot(aes(x = time, y = Ieq.structure, color = disturbance)) +
  geom_point() +
  xlab('Time (years)') +
  ylab('Cummulative integral of\ndistance to forest structure equilibrium\n(AGB, BA, N, N10 and N30)') +
  scale_color_manual('Disturbance\nintensity\n(% of BA)',
                     values = cols)
cowplot::plot_grid(g_agb, g_resilience, g_deq, g_Ieq, labels = LETTERS[1:4])
```

@@

The effect of biodiversity on ecosystem resilience can be split in complementarity and selection effect using @Loreau2001 partitioning:

@@

$X_{sp}$ is the ecosystem variable value for one species either in mixture $X_{sp}(mixture)$ or in monoculture $X_{sp}(monoculture)$. $P_{sp}$ is the proportion of the species in the mixture represented by species relative abundance. Consequently, $CE$ averages diversity effects of all species presents in the mixture (both negatives and positives). Whereas $SE$ become positive when dominant species outperform themselves in mixture than in monoculture, and negative when less dominant species outperform themselves in mixture than in monoculture [@Tobner2016]. But similarly, to resilience measurement, biodiversity net effect $NE$ is a dynamic equilibrium and vary over time without disturbance. So in order to correctly assess selection and complementarity effect in answer to disturbance, we normalized it by undisturbed control ecosystem net effect $NE_C$ to measure treatments net effect resilience $R(NE_T)$:

@@

Resilience trajectories of ecosystem variable after disturbance were partitioned between complementarity effect $CE$ and selection effect $SE$. In order to do that, we simulated the 163 species individually representing 652 simulations of monoculture.

@@

Species richness was not directly link to $Ieq_{600}$. Increased functional diversity reduced cumulative integral from ecosystem distance to equilibrium after 600 years ($Ieq_{600}$). In addition, functional evenness was complementary reducing $Ieq_{600}$. Finally, low species richness could result in variant $Ieq_{600}$, but increased species richness resulted in increased functional diversity and consequently lower $Ieq_{600}$. We found similar results for all disturbance levels and forest functioning (see [Appendix 5: Disturbance simulations]).

@@

The net biodiversity effect was partitioned between selection and complementarity effect after normalizing using undisturbed control simulation (see Table \@ref(tab:tNE)) to measure the resilience of each ecosystem measure, e.g. aboveground biomass (see Figure \@ref(fig:gBE)). We found that complementarity effect was dominant in the biodiversity net effect in the first decades. But after few decades the complementarity effect diminished toward an extremely low value. On the contrary, selection effect was reduced in the initial times after disturbance. It increased during the first decades and was then greater. The time lag for which complementarity effect overcame the selection effect increased with disturbance intensity. Finally, 600 years after disturbance, the biodiversity net effect on aboveground biomass was still not at equilibrium for a disturbance intensity greater than 25% of basal area. We found similar results but with an amplified signal for basal area ($BA$) and stem abundance ($N$ but with an inverted signal because of the forest self thinning) (see Supplementary materials S2). Finally, forest growth primary productivity ($GPP$) recovered in few years (proportionally to disturbance), and its net effect was maintained by complementarity effect (see Supplementary materials S2).

@@

In the limit of the model, we were able to show that functional diversity improved tropical forest resilience after disturbance. More particularly, functional diversity and evenness are key components of diversity in forest recovery trajectory after disturbance. Moreover, we found that complementarity between species was insuring forest recovery in the early times before more productive species dominate the forest and insure late recovery. We thus advocate the view that both diversity theory (with niche complementarity and facilitation) and mas ratio theory (represented by selection) are implied in ecosystem resilience but with relative importance depending on the spatial grain and time since the disturbance event.

## Functional diversity improve tropical forest resilience
